The weighted hyperdirichlet distribution: the `hyper3` class

### 2024-01-10 inst/hyper3.Rmd
consistentification
typos

---
title: "The weighted hyperdirichlet distribution: the `hyper3` class"
author: "R. K. S. Hankin"
output: bookdown::html_document2
bibliography: jss_hyper3.bib
---

@@

To cite the `hyper2` package in publications, please use @hankin2017a.
Objects of class \code{hyper3} are a generalization of \code{hyper2}
objects that allow the brackets to contain weighted probabilities.
Likelihood functions are defined on non-negative $p_1,\ldots, p_n$
subject to the unit-sum constraint $\sum p_i=1$.  Given known weights
$w^i_j$ with $1\leq i\leq j$ we have

@@

$$\mathcal{L}(p_1,p_2,\lambda;A,B,C,D)=
\left(\frac{\lambda p_1}{\lambda p_1 + p_2}\right)^{A}
\left(\frac{p_2        }{\lambda p_1 + p_2}\right)^{B}
\left(\frac{p_1        }{p_1 + \lambda p_2}\right)^{C}
\left(\frac{\lambda p_2}{p_1 + \lambda p_2}\right)^{D}.
$$

@@

$$
\hat{p_1}=\frac{\sqrt{AC}}{\sqrt{AC}+\sqrt{BD}}\qquad
\hat{p_2}=\frac{\sqrt{BD}}{\sqrt{AC}+\sqrt{BD}}\qquad
\hat{\lambda}=\sqrt{\frac{AD}{BC}}
$$

If $A=x,B=x,C=x,D=x$ and $\lambda=1.3$ [assumed for the moment to be
known], appropriate package idiom might be:

```{r,label=usenamedvectorsforH}
library("hyper2",quietly=TRUE)
A <- 6 ; B <- 2 ; C <- 7 ; D <- 1
H <- hyper3()
H[c(p1=1.3)]      %<>% inc(A) 
H[c(p2=1)]        %<>% inc(B) 
H[c(p1=1.3,p2=1)] %<>% dec(A+B)
H[c(p1=1)]        %<>% inc(C) 
H[c(p2=1.3)]      %<>% inc(D) 
H[c(p1=1,p2=1.3)] %<>% dec(C+D)
H
```

@@

Above, obect `M` has real parts being home wins and imaginary parts
being away wins.

We may estimate $p_1$ and $p_2$ using maximum likelihood, `maxp()` in
package idiom:

@@

Another motivating example might come from Plackett-Luce likelihood
functions for order statistics.  Suppose we have three players with
nonnegative strengths $p_\mathrm{M},p_\mathrm{RB},p_\mathrm{F}$, with
unit sum.  These players might be conceputalised as F1 constructors
(here Mercedes (M), Red Bull (RB), and Ferrari (F) for the sake of
argument).  Each constructor fields two cars and the order statistic
might be

>>>>

---
title: "The weighted hyperdirichlet distribution: the `hyper3` class"
author: "R. K. S. Hankin"
output: bookdown::html_document2
bibliography: hyper2.bib
---

@@

To cite the `hyper2` package in publications, please use @hankin2017.
Objects of class `hyper3` are a generalization of `hyper2`
objects that allow the brackets to contain weighted probabilities.
Likelihood functions are defined on non-negative numbers $p_1,\ldots, p_n$
subject to the unit-sum constraint $\sum p_i=1$.  Given known weights
$w^i_j$ with $1\leq i\leq j$ we have

@@

$$\mathcal{L}(p_1,p_2,\lambda;a,b,c,d)=
\left(\frac{\lambda p_1}{\lambda p_1 + p_2}\right)^{a}
\left(\frac{p_2        }{\lambda p_1 + p_2}\right)^{b}
\left(\frac{p_1        }{p_1 + \lambda p_2}\right)^{c}
\left(\frac{\lambda p_2}{p_1 + \lambda p_2}\right)^{d}.
$$

@@

$$
\hat{p_1}=\frac{\sqrt{ac}}{\sqrt{ac}+\sqrt{bd}}\qquad
\hat{p_2}=\frac{\sqrt{bd}}{\sqrt{ac}+\sqrt{bd}}\qquad
\hat{\lambda}=\sqrt{\frac{ad}{bc}}
$$
Likelihoods of this form are easily specified using package idiom.

If $a=6,b=2,c=7,d=1$ and $\lambda=1.3$ [assumed for the moment to be
known], appropriate package idiom might be:

```{r,label=usenamedvectorsforH}
library("hyper2",quietly=TRUE)
A <- 6 ; B <- 2 ; C <- 7 ; D <- 1 # upper-case to avoid conflicts
H <- hyper3()
H[c(p1=1.3)]      %<>% inc(A) 
H[c(p2=1)]        %<>% inc(B) 
H[c(p1=1.3,p2=1)] %<>% dec(A+B)
H[c(p1=1)]        %<>% inc(C) 
H[c(p2=1.3)]      %<>% inc(D) 
H[c(p1=1,p2=1.3)] %<>% dec(C+D)
H
```

@@

Above, object `M` is a $2\times 2$ complex matrix, with real parts being home
wins and imaginary parts being away wins.  See how `home_away3(M)` returns the
same likelihood function as the explicit calculation performed by hand
previously.  Now, we can calculate maximum likelihood estimates
$\hat{p_1},\hat{p_2}$ of $p_1$ and $p_2$ using numerical optimization
techniques, `maxp()` in package idiom:

@@

Another motivating example might come from Plackett-Luce likelihood
functions for order statistics.  Suppose we have three players with
nonnegative strengths $p_\mathrm{M},p_\mathrm{RB},p_\mathrm{F}$, with
unit sum.  These players might be conceptualised as F1 constructors
(here Mercedes (M), Red Bull (RB), and Ferrari (F) for the sake of
argument).  Each constructor fields two cars and the order statistic
might be
